Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-03-14 paper/paper1/ms.tex
Make a Matrix equation out of Voigt Profile figures, start telluric section

\section{Methodology}\label{methodology}

@@

\subsection{The \emph{blas\'e} clone model}

@@

\begin{equation}
    \begin{pmatrix}
        1 -\mathsf{V}_1(\bm{\lambda}_s)                     & \\
        1 - \mathsf{V}_2(\bm{\lambda}_s)                    & \\
        \vdots                                              & \\
        1 - \mathsf{V}_{N_{\mathrm{lines}}}(\bm{\lambda}_s) &
    \end{pmatrix}
\end{equation}

An element of this matrix, $F_{j\lambda}$, will have the flux value for a given line at a given wavelength coordinate. Equation \ref{equation1} performs a type of matrix contraction, turning a $N_{\mathrm{lines}}\times N_{S}$ matrix into a length $N_{S}$ row vector. The number of Floating Point Operations (FLOPS) scales with the number of entries in this matrix. We can rewrite Equation \ref{equation1} as a sum by taking the log of both sides and dropping in this $\bm{\bar{F}}$ matrix:

\begin{eqnarray}
    \ln{\mathsf{S}_{\rm clone}} = \sum_{j=1}^{N_{lines}} \ln{F_{j\lambda}} = \mathbf{1} \cdot \ln{\bm{\bar{F}}}
\end{eqnarray}

where $\mathbf{1}$ is a $1\times N_{\rm lines}$ row vector. Efficient GPU algorithms exist for matrix manipulations such as this one, allowing this computation to proceed quickly on modern machines. In particular, the proprietary CUDA architecture for NVIDIA\textsuperscript{\tiny\textregistered} GPUs contains Tensor cores with specialized matrix math. The chief bottleneck occurs when the storage of the $\bm{\bar{V}}$ matrix exceeds the available RAM of a GPU or CPU: the computation will fail with an ``Out of Memory'' exception. Modern NVIDIA GPUs have $8-40$ GB of RAM. The memory bottleneck is even more pernicious than mere storage of the $\bm{\bar{V}}$ matrix since the CPU/GPU also has to store the computation graph for autodiff. It is generally not possible to evaluate Equation \ref{equation1} in its entirety in one-fell-swoop. A remedy is needed.

@@

We retain a relatively small number of pixels $N_{\rm cut}$ adjacent to the line center. Setting this wing cut produces a speedup by a factor of $\frac{N_S}{N_{\mathrm{cut}}}$, which can exceed $100\times$ for wide bandwidth spectra.The choice of $N_{\rm cut}$ is nuanced.  It should be set large enough that truncation effects are not seen for the broadest lines.  But even more, $N_{\rm cut}$ has to be future-proofed for Doppler-shifting. Extreme Doppler shifts could hypothetically send line cores entirely outside the extents of $N_{\rm cut}$.  We therefore typically set wingcuts comparable to the buffer size $\pm \Delta \lambda_{\mathrm{buffer}}$, even though most weak lines only perceptibly affect $<1\;$\AA. We coerce all wing cuts to be the same number of pixels, typically 6000 pixels for \texttt{PHOENIX}, with the middle pixel being at the line center position, and about 3000 pixels to the red and blue side of the line. We refer to this approximate sparse matrix as $\ln{\bm{\hat{F}}}$.

The PHOENIX spectra double their pixel sampling shortward of $\lambda = 1 \mu$m, so 6000 pixels in the visible is about 30 Angstroms, but 6000 pixels in the infrared is about 60 \AA. We \emph{could have} demanded that the wing cut be exactly 30 \AA, or some other heuristic, but that choice would result in so-called \emph{jagged} or \emph{ragged} arrays: collections of 1D arrays that differ in their lengths. Some lines would have 5999 pixels, some would have 6000, and others would have 2999 or 3000 pixels, depending whether they occur to the red or blue side of the $1 \;\mu$m kink in sampling. Most efficient CPU and GPU algorithms are not equipped to deal with ragged arrays, therefore increasing the computation time. Instead, the standardization choice of exactly 6000 pixels means we can use standard matrix or tensor arithmetic, which somewhat counterintuitively is faster while handling more pixels for all spectral lines: 6000 pixels everywhere is faster than 6000 pixels for some and 2999 for other lines.  Locating the indices of these 6000 ``nearest-neighbor'' pixels is very fast, only occurs once, and then is fixed.

@@

We use the Adam optimizer \citep{2014arXiv1412.6980K} with a typical learning rate $LR\in (0.005, 0.05)$ and all the defaults for \texttt{PyTorch v1.11}.  We defined the number of training epochs $N_{epoch}=100-10,000$ depending on the application. The user can optionally monitor a live view of the training progress with Tensorboard\authorcomment1{citation} to gain an intuition for the training efficiency.

@@

Figure \ref{fig_cloned_spectrum_demo} shows a portion of a PHOENIX spectrum cloned with \emph{blas\'e}. The $1000$ epochs of training took 56 seconds on an NVIDIA\textsuperscript{\tiny\textregistered} RTX2070 GPU with \texttt{PyTorch v1.11}, \texttt{CUDA v11.1}, and Intel\textsuperscript{\tiny\textregistered} Core\textsuperscript{\tiny TM} i7-9750H CPUs at 2.60GHz, with all tensors as FP64. The same computation on a 2020 M1 Macbook Air took 1$^h$25$^m$ with \texttt{PyTorch v1.9}, $90\times$ slower than the GPU counterpart.

We store the model parameters to disk and refer to the entire collection of parameters as the \emph{pre-trained model}.

@@

We compute the residual $\mathsf{R}(\bm{\lambda}_s)$ of native PHOENIX minus cloned model, illustrated in the bottom panel of Figure \ref{fig_cloned_spectrum_demo}. We see an RMS residuals of 1.2\%/pixel at native resolution. We see three main categories of cloning flaws.

@@

\begin{eqnarray}
    \mathsf{F}_{ex}(\bm{\lambda}) = \mathsf{F}(\bm{v}-RV) * \zeta \left(\frac{\bm{v}}{v\sin{i}}\right) \label{eqn_convolution}
\end{eqnarray}

@@

Operationally, the Doppler shift gets applied to the line center positions $\tilde \lambda_{c,i} = (\lambda_{c,i}-\delta \lambda)$, and not the individual wavelength grid point coordinates, $\bm{\lambda}$. This choice yields a key benefit: it makes the $RV$ autodiff-aware, meaning that an infinitesimal change to the RV value can easily be sensed through backpropagation.  The strategy allows relatively low computational and memory cost since it comports with the sparse matrix implementation.

The instrumental resolving power, $R$, acts as a convolution with a Gaussian line profile of width $\sigma=\frac{c}{2.355 R}$.  Real astronomical instruments usually have wavelength-dependent resolving power, which complicates the implementation.  The extent to which this effect matters will depend on the science case.  For now, \emph{blas\'e} simply assumes a fixed resolving power.

>>>>

\section{Methodology I: Cloning stellar spectra}\label{methodology}

@@

\subsection{The \texttt{blas\'e} Stellar Clone Model}

@@

\begin{equation}
    \begin{pmatrix}
        1 - a_1 \mathsf{V}_1(\bm{\lambda}_s)                     & \\
        1 - a_2 \mathsf{V}_2(\bm{\lambda}_s)                    & \\
        \vdots                                              & \\
        1 - a_{N_{\mathrm{lines}}} \mathsf{V}_{N_{\mathrm{lines}}}(\bm{\lambda}_s) &
    \end{pmatrix}
\end{equation}

An element of this matrix, $F_{ji}$, will have the flux value for a given line at a given wavelength coordinate. Equation \ref{equation1} performs a type of matrix contraction, turning a $N_{\mathrm{lines}}\times N_{S}$ matrix into a length $N_{S}$ row vector. The number of Floating Point Operations (FLOPS) scales with the number of entries in this matrix. We can rewrite Equation \ref{equation1} as a sum by taking the log of both sides and dropping in this $\bm{\bar{F}}$ matrix:

\begin{eqnarray}
    \ln{\mathsf{S}_{\rm clone}} = \sum_{j=1}^{N_{lines}} \ln{F_{ji}} = \mathbf{1} \cdot \ln{\bm{\bar{F}}}
\end{eqnarray}

where $\mathbf{1}$ is a $1\times N_{\rm lines}$ row vector of all-ones. We re\"emphasize that each spectral line has to be painstakingly evaluated across the entire spectral bandwidth.   Efficient GPU algorithms exist for voluminous matrix manipulations such as this one, so this computation can proceed as quickly as possible on modern machines. In particular, the proprietary CUDA architecture for NVIDIA\textsuperscript{\tiny\textregistered} GPUs contains Tensor cores with specialized matrix math. The chief bottleneck occurs when the storage of the $\bm{\bar{F}}$ matrix exceeds the available RAM of a GPU or CPU: the computation will fail with an ``Out of Memory'' exception. Modern NVIDIA GPUs have $8-40$ GB of RAM, which translates roughly to a few thousand spectral lines across $\sim$300,000 pixels.  The memory bottleneck is even more pernicious than mere storage of the $\bm{\bar{F}}$ matrix since the CPU/GPU also has to store the computation graph for autodiff. It is generally not possible to evaluate Equation \ref{equation1} in its entirety in one-fell-swoop. A remedy is needed.

@@

We retain a relatively small number of pixels $N_{\rm cut}$ adjacent to the line center. Setting this wing cut produces a speedup by a factor of $\frac{N_S}{N_{\mathrm{cut}}}$, which can exceed $100\times$ for wide bandwidth spectra.The choice of $N_{\rm cut}$ is nuanced.  It should be set large enough that truncation effects are not seen for the broadest lines.  But even more, $N_{\rm cut}$ has to be future-proofed for Doppler-shifting. Extreme Doppler shifts could hypothetically send line cores entirely outside the extents of $N_{\rm cut}$ if set too low.  We therefore typically set wingcuts comparable to the buffer size $\pm \Delta \lambda_{\mathrm{buffer}}$, even though most weak lines only perceptibly affect $<1\;$\AA. We coerce all wing cuts to be the same number of pixels, typically 6000 pixels for \texttt{PHOENIX}, with the middle pixel being at the line center position, and about 3000 pixels to the red and blue side of the line. We populate a new approximate sparse matrix $\ln{\bm{\hat{F}}}$ with only these $~6000$ pixels per line and assume zeros everywhere else.
%\footnote{The PHOENIX spectra double their pixel sampling shortward of $\lambda = 1 \mu$m, so 6000 pixels in the visible is about 30 Angstroms, but 6000 pixels in the infrared is about 60 \AA. We \emph{could have} demanded that the wing cut be exactly 30 \AA, or some other heuristic, but that choice would result in so-called \emph{jagged} or \emph{ragged} arrays: collections of 1D arrays that differ in their lengths. Some lines would have 5999 pixels, some would have 6000, and others would have 2999 or 3000 pixels, depending whether they occur to the red or blue side of the $1 \;\mu$m kink in sampling. Most efficient CPU and GPU algorithms are not equipped to deal with ragged arrays, therefore increasing the computation time. Instead, the standardization choice of exactly 6000 pixels means we can use standard matrix or tensor arithmetic, which somewhat counterintuitively is faster while handling more pixels for all spectral lines: 6000 pixels everywhere is faster than 6000 pixels for some and 2999 for other lines.  Locating the indices of these 6000 ``nearest-neighbor'' pixels is very fast, only occurs once, and then is fixed.}

The remapping of the sparse matrix can be visualized as the shifting of all lines to the center of this new matrix $\bm{\hat{F}}$.  The algorithmic machinery keeps track of each $(i, j, F_{ji})$ trio of coordinates and flux values.

\begin{equation} \bm{\bar{F}} = 
    \begin{pmatrix}
         \cdots \includegraphics[height=1cm]{figures/voigt_01.pdf}  \cdots & \\
         \cdots \includegraphics[height=1cm]{figures/voigt_02.pdf}  \cdots & \\
        \vdots                                              & \\
        \cdots \includegraphics[height=1cm]{figures/voigt_03.pdf} \cdots &
    \end{pmatrix} \mapsto 
    \begin{pmatrix}
        \includegraphics[height=1cm]{figures/voigt_01b.pdf}  & \\
        \includegraphics[height=1cm]{figures/voigt_02b.pdf}  & \\
       \vdots                                              & \\
        \includegraphics[height=1cm]{figures/voigt_03b.pdf} &
   \end{pmatrix} = \bm{\hat{F}}
\end{equation}

@@

We use the Adam optimizer \citep{2014arXiv1412.6980K} with a typical learning rate $LR\in (0.005, 0.05)$ and all the defaults for \texttt{PyTorch v1.11}.  We defined the number of training epochs $N_{epoch}=100-10,000$ depending on the application. The user can optionally monitor a live view of the training progress with Tensorboard \citep{tensorflow2015-whitepaper} to gain an intuition for the training efficiency.

@@

Figure \ref{fig_cloned_spectrum_demo} shows a portion of a PHOENIX spectrum cloned with \texttt{blas\'e}. The $1000$ epochs of training took 56 seconds on an NVIDIA\textsuperscript{\tiny\textregistered} RTX2070 GPU with \texttt{PyTorch v1.11}, \texttt{CUDA v11.1}, and Intel\textsuperscript{\tiny\textregistered} Core\textsuperscript{\tiny TM} i7-9750H CPUs at 2.60GHz, with all tensors as FP64. The same computation on a 2020 M1 Macbook Air took 1$^h$25$^m$ with \texttt{PyTorch v1.9}, $90\times$ slower than the GPU counterpart.

We store the model parameters to disk and refer to the entire collection of parameters as a \emph{pre-trained model}.  More specifically this fine-tuned model represents an evaluable and interpretable clone of the original static pixel values flux values.

\section{Methodology II: Cloning Telluric Spectra}
Ground-based near-IR \'echelle spectra possess countless depressions attributable to molecular line absorption in Earth's atmosphere.  These telluric lines hamper the unbiased interpretation of \'echelle spectra, so some treatment plan is needed.  Often the regions of known, deep tellurics are simply discarded.  In other cases, the lines are modeled with first principles line-by-line radiative transfer \citep[\emph{e.g.} \texttt{TelFit,}][]{2014AJ....148...53G} or through data-driven means \citep[\emph{e.g.} \texttt{wobble,}][]{2019AJ....158..164B}.  The most demanding EPRV applications require a precision characterization of telluric lines that the astronomical community has not yet been able to achieve, and that may rival even the abilities of Earth Science practitioners.  A hybrid data-/model- driven approach was the chief recommendation of the \emph{Telluric Hack Week} workshop aimed at improving our mitigation of the atmosphere's deleterious effects.  The \texttt{blas\'e} framework achieves a key milestone by introducing this \emph{hybrid} data- and model-driven approach.

\subsection{Adjustments from the stellar model}
The Telluric clone model proceeds almost identically to the stellar model, with the following modifications.

@@

We compute the residual $\mathsf{R}(\bm{\lambda}_S)$ of native PHOENIX minus cloned model, illustrated in the bottom panel of Figure \ref{fig_cloned_spectrum_demo}. We see an RMS residuals of 1.2\%/pixel at native resolution. We identify three main categories of cloning flaws.

@@

\begin{eqnarray}
    \mathsf{S}_{\rm ext}(\bm{\lambda}_S) = \mathsf{S}_{\rm clone}(\bm{v}-RV) * \zeta \left(\frac{\bm{v}}{v\sin{i}}\right) \label{eqn_convolution}
\end{eqnarray}

@@

Operationally, the Doppler shift gets applied to the line center positions $\tilde \lambda_{c,j} = (\lambda_{c,j}-\delta \lambda)$, and not the individual wavelength grid point coordinates, $\bm{\lambda}_S$. This choice yields a key benefit: it makes the $RV$ autodiff-aware, meaning that an infinitesimal change to the RV value can easily be sensed through backpropagation by affecting only the line center positions.  The strategy allows relatively low computational and memory cost since it comports with the sparse matrix implementation.

The instrumental resolving power, $R$, acts as a convolution with a Gaussian line profile of width $\sigma=\frac{c}{2.355 R}$.  Real astronomical instruments usually have wavelength-dependent resolving power, which complicates the implementation.  The extent to which this effect matters will depend on the science case.  For now, \texttt{blas\'e} simply assumes a fixed resolving power.
